Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-03 ind_files/01_Introduction.tex
Minor phrasing changes, change epigraph from Potochnik to Harper

\epigraph{\justifying The epistemic aim of science is not truth, but understanding}{Angela Potochnik}

More than 150 years have passed since Ernst Haeckel first coined the term ‘ecology’ in 1866. Today, ecology and evolution are incredibly interdisciplinary, borrowing techniques and ideas from diverse fields such as computer science, statistics, economics, dynamical systems, physics, and information theory. With this development has come a cornucopia of models that use these tools and techniques to try and understand biological phenomena. Though many of these models deal with specific `low-level' factors, there is also value to formulating general, idealized models in abstract terms to underscore these organizing principles and better illustrate the fundamental processes that are required to capture the `essence' of a biological pattern \citep{frank_natural_2012, vellend_theory_2016}. Such abstractions and idealizations are useful in unifying apparently disparate modelling frameworks and are thus vital to theory-building \citep{luque_mirror_2021}.

@@

The push and pull between the search for general patterns and the specification of minutiae has a long and torturous history in ecology \citep{kingsland_modeling_1985}. One of the first mathematical idealizations in ecology came in the form of the logistic equation formulated by Pearl and Verhulst, and shortly thereafter, equations for the populations of interacting predator and prey species, put forth independently by Lotka and Volterra. These models were immediately controversial, and for good reason: Many ecologists felt that they were overly idealistic and neglected many important truths about real biological populations \citep{kingsland_modeling_1985}. These models were nevertheless quite good at predicting the patterns of such populations, and have proved valuable to the field of population ecology. Today, these models are viewed as `classical' and are regularly used even by hard-line empiricists, not because we believe them to be true in all their gory biological details, but because we recognize that they can be \emph{useful} despite being blatantly false generalizations. This is a single instance of a more general philosophical idea concerning the goals and ideals of science. As suggested by the epigraph at the beginning of the chapter, philosophers of science \citep{potochnik_idealization_2018} have recently argued that science does not seek truth, but instead seeks understanding. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses \citep{zuk_models_2018}. Relatively simple models of complex eco-evolutionary processes are therefore desirable as a way to shine light on these phenomena. To reiterate once more, the goal of such simple models is not truth (whatever that is), predictive power (as with models in physics), or detailed description (as with detailed individual-based simulations, very flexible statistical models, or machine learning), but \emph{understanding}. Since the world is complicated and humans are limited, such understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. It is important to remember at the outset that the models I speak about in this thesis may seem to be over-idealized and too general, and will make only qualitative predictions. This is \textit{by design}, in pursuit of general insight over precise quantitative prediction.\\
Vellend has recently argued that conceptual synthesis in community ecology requires ``shifting the emphasis away from an organizational structure based on the useful lines of inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns" \citep{vellend_theory_2016}. Vellend's assertion is based on the fact that population genetics has managed to come up with reasonably comprehensive theory due to its focus on the abstract `high-level' processes of selection, mutation, drift, and gene flow instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on specific `low-level' processes such as predation rate, limiting resources ($R^*$), storage effects, priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table 5.1 in \citep{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that community ecology `is a mess'. Vellend proposes organizing ecological models according to the `high-level' processes of selection, ecological drift (demographic stochasticity), dispersal, and speciation.\\ Of course, no such general organization will be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”\citep{kingsland_modeling_1985}. However, I believe the act of looking beyond the low-level processes present in biological systems and categorizing theories, models, and concepts in terms of a small number of (blurry-eyed) fundamental high-level processes provides a powerful unifying tool to organize concepts in biology. This is perhaps best exemplified by Darwin’s theory of natural selection as first proposed in \emph{The Origin of Species}. Darwin famously painstakingly collected a series of ‘low-level’ facts and observations regarding breeds, wild populations, and the geographic record to support his hypothesis. However, ultimately, these observations culminated in a synthesis whereby they were all unified under a single, abstract, ‘high-level’ process, namely evolution by natural selection. A similar focus on high-level processes is also present in the mathematization of evolution by natural selection carried out by evolutionary biologists during and after the Modern Synthesis.

@@

Classical population genetics regarded forces like selection and mutation as fundamental, and the success of this approach during the Modern Synthesis illustrates the value of formulating high-level, abstract models that only provide a `high-level' description of the fundamental processes required to capture the essence of a biological pattern. However, the evolutionary play famously unfolds in the ecological theatre \citep{hutchinson_ecological_1965}. Thus, quantities like fitness are not truly fundamental but instead emerge as the net result of various ecological interactions, tradeoffs, and constraints, a fact that can have important consequences for evolution \citep{coulson_putting_2006, kokko_can_2017}. Trying to model such `eco-evolutionary dynamics' has sprouted a rich body of theoretical literature and led to the development of theoretical frameworks like evolutionary game theory and adaptive dynamics, fields which have greatly enriched our understanding of biological populations \citep{brown_why_2016}. Eco-evolutionary population dynamics can broadly be organized under a single unifying framework, the Price equation, that yields all the relevant formal structures under various limits \citep{page_unifying_2002, lion_theoretical_2018}. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of the high-level evolutionary forces of selection, mutation, and drift, thus providing a useful conceptual framework for thinking about how populations change over time \citep{frank_natural_2012}. The Price equation also leads to a small number of simple yet insightful `fundamental theorems' of population biology \citep{queller_fundamental_2017, lion_theoretical_2018, lehtonen_price_2018} and unifies several various seemingly disjoint formal structures under a single theoretical banner \citep{ lehtonen_price_2020, luque_mirror_2021}\\

@@

Incorporating stochasticity into deterministic systems is a tricky business, and, if done in a phenomenological manner by adding noise to a `deterministic skeleton' \citep{coulson_skeletons_2004} in an ad-hoc fashion, can lead to nonsensical predictions and inconsistencies \citep{strang_how_2019}. Stochastic individual-based models, in which (probabilistic) rules are specified at the level of the individual and population level dynamics are systematically derived from first principles, are much more natural \citep{black_stochastic_2012}, lead to sensible predictions \citep{strang_how_2019}, and can fundamentally differ from the predictions made by simply adding noise terms to a deterministic model \citep{strang_how_2019} (This is also true when incorporating spatial structure into non-spatial models, see for example \cite{durrett_importance_1994}). Formulating the fundamental formal structures of evolutionary biology in terms of the mechanistic demographic processes of birth and death at the individual level is also greatly desirable for biological reasons \citep{metcalf_why_2007,geritz_mathematical_2012} because `all paths to fitness lead through demography' \citep{metcalf_all_2007}. In other words, since demographic processes such as birth and death rates explicitly account for the ecology of the system, they can more accurately reflect the complex interplay between ecological and evolutionary processes and provide a more fundamental mechanistic description of the relevant evolutionary forces and population dynamics \citep{doebeli_towards_2017}.

@@

In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. Part \ref{part_theory} presents the mathematical formalism in a detailed, self-contained, pedagogical manner. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, use the system size expansion to arrive at a continuous description of change in trait frequencies as an SDE. Under mild assumptions, I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations from stochastic first principles. While the mathematics of section \ref{sec_nD_processes} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects. As such, chapter \ref{sec_nD_processes} can be thought of as a tutorial, conceptual synthesis, and unification. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck in terms of functional derivatives. Under mild assumptions, I show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues \citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, is entirely original both mathematically and biologically. Mathematically, the ideas present heuristic, accessible alternatives to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits \citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021} in terms of a `stochastic field theory'. This formulation in terms of stochastic field equations also opens up the model to analysis using the powerful heuristic tools of physics such as the path integral \citep{doi_second_1976, peliti_path_1985, dodd_many-body_2009, chow_path_2015, weber_master_2017}, though I do not attempt to leverage these connections here, leaving it as an opportunity for future study. Examples for one-dimensional, multi-dimensional, and quantitative trait models that can be analyzed in this formalism are presented in Appendix \ref{App_examples}.\\

>>>>

\epigraph{\justifying The theory of evolution by natural selection is an ecological theory—founded on ecological observation by perhaps the greatest of all ecologists. It has been adopted by and brought up by the science of genetics, and ecologists, being modest people, are apt to forget their distinguished parenthood}{John Harper}

More than 150 years have passed since Charles Darwin first published \textit{Origin} (in 1859) and Ernst Haeckel first coined the term ‘ecology’ (in 1866). Today, both ecology and evolution are incredibly interdisciplinary fields, borrowing techniques and ideas from diverse areas such as computer science, statistics, economics, dynamical systems, physics, and information theory. With this development has come a cornucopia of models that try to understand biological phenomena in the language of these borrowed tools and techniques. Though many of these ideas are used to accurately describe and analyze specific systems, there is also value to formulating general models in abstract terms that only incorporate a small number of `fundamental' processes and try to capture the `essence' of a biological pattern \citep{frank_natural_2012, vellend_theory_2016,luque_mirror_2021}. Such general organizing models are vital to theory-building \citep{luque_mirror_2021} because they help clarify conceptual similarities and unifying factors between apparently disparate modelling frameworks, helping us seamlessly translate essential ideas from one theoretical language to another.

@@

The push and pull between the search for general patterns and the specification of minutiae has a long and torturous history in ecology \citep{kingsland_modeling_1985} and evolution \citep{provine_origins_2001}. One of the first mathematical idealizations in ecology came in the form of the logistic equation formulated by Pearl and Verhulst, and shortly thereafter, equations for the populations of interacting predator and prey species, put forth independently by Lotka and Volterra. These models were immediately controversial, and for good reason: Many ecologists felt that they were overly idealistic and neglected many important truths about real biological populations \citep{kingsland_modeling_1985}. These models were nevertheless quite good at predicting the patterns of such populations, and have proved valuable to the field of population ecology. Today, these models are viewed as `classical' and are regularly used even by hard-line empiricists, not because we believe them to be true in all their gory biological details, but because we recognize that they can be \emph{useful} despite being blatantly false generalizations. This is a single instance of a more general philosophical idea concerning the goals and ideals of science. As suggested by the epigraph at the beginning of the chapter, philosophers of science \citep{potochnik_idealization_2018} have recently argued that science does not seek truth, but instead seeks understanding. The fact that idealization is part and parcel of science is clear if one looks at the actual practice, be it theorists making unrealistic assumptions on paper to model specific phenomena or experimentalists creating artificially controlled conditions in the laboratory to test specific hypotheses \citep{zuk_models_2018}. Relatively simple models of complex eco-evolutionary processes are therefore desirable as a way to shine light on these phenomena. To reiterate once more, the goal of such simple models is not truth (whatever that is), predictive power (as with models in physics), or detailed description (as with detailed individual-based simulations, very flexible statistical models, or machine learning), but \emph{understanding}. Since the world is complicated and humans are limited, such understanding inevitably comes at the cost of other desirable qualities such as the ability to make precise quantitative predictions. It is important to remember at the outset that the models I speak about in this thesis may seem to be too general and may seem to ignore many biological details. This is \textit{by design}, in pursuit of general insight over precise quantitative prediction.\\
Vellend has recently argued that conceptual synthesis in community ecology requires ``shifting the emphasis away from an organizational structure based on the useful lines of inquiry carved out by researchers, to one based on the fundamental processes that underlie community dynamics and patterns" \citep{vellend_theory_2016}. Vellend's assertion is based on the fact that population genetics has managed to come up with reasonably comprehensive theory due to its focus on the abstract `high-level' processes of selection, mutation, drift, and gene flow instead of the myriad `low-level' processes that may be responsible for generating them. In contrast, he believes that practitioners of community ecology often focus on specific `low-level' processes such as predation rate, limiting resources ($R^*$), storage effects, priority effects, senescence, and niche partitioning, leading to a plethora of models (see Table 5.1 in \citep{vellend_theory_2016} for an in-exhaustive list of 24 such models) and the conclusion that community ecology `is a mess'. Vellend proposes organizing ecological models according to the `high-level' processes of selection, ecological drift (demographic stochasticity), dispersal, and speciation.\\ Of course, no such general organization will be perfect or all-encompassing. As Robert MacArthur once remarked, ``general events are only seen by ecologists with rather blurred vision. The very sharp-sighted always find discrepancies and are able to see that there is no generality, only a spectrum of special cases”\citep{kingsland_modeling_1985}. However, I believe the act of looking beyond the low-level processes present in biological systems and categorizing theories, models, and concepts in terms of a small number of (blurry-eyed) fundamental high-level processes provides a powerful unifying tool to organize concepts in biology.

@@

Classical population genetics regarded forces like selection and mutation as fundamental, and the success of this approach during the Modern Synthesis illustrates the value of formulating high-level, abstract models that only provide a `high-level' description of the fundamental processes required to capture the essence of a biological pattern. However, the evolutionary play famously unfolds in the ecological theatre \citep{hutchinson_ecological_1965}. Thus, quantities like fitness are not truly fundamental but instead emerge as the net result of various ecological interactions, tradeoffs, and constraints \citep{metz_how_1992}, a fact that can have important consequences for evolution \citep{coulson_putting_2006, kokko_can_2017}. Trying to understand such `eco-evolutionary dynamics' has sprouted a rich body of theoretical literature under the broad heading of `evolutionary ecology' and led to the development of theoretical frameworks like evolutionary game theory and adaptive dynamics, fields which have greatly enriched our understanding of biological populations \citep{brown_why_2016}. Eco-evolutionary population dynamics can (very) broadly be organized under a single unifying framework, the Price equation, that yields all the relevant formal structures (such as evolutionary game theory and classic population genetics) as special cases \citep{page_unifying_2002, queller_fundamental_2017, lion_theoretical_2018}. The Price equation partitions changes in population composition into multiple terms, each of which lends itself to a straightforward interpretation in terms of the high-level evolutionary forces of selection, mutation, and drift, thus providing a useful conceptual framework for thinking about how populations change over time \citep{frank_natural_2012}. The Price equation also leads to a small number of simple yet insightful `fundamental theorems' of population biology \citep{queller_fundamental_2017, lion_theoretical_2018, lehtonen_price_2018} and unifies several various seemingly disjoint formal structures under a single theoretical banner \citep{ lehtonen_price_2020, luque_mirror_2021}\\

@@

Incorporating stochasticity into deterministic systems is a tricky business, and, if done in a phenomenological manner by adding noise to a `deterministic skeleton' \citep{coulson_skeletons_2004} in an ad-hoc fashion, can lead to nonsensical predictions and inconsistencies \citep{strang_how_2019}. Stochastic individual-based models, in which (probabilistic) rules are specified at the level of the individual and population level dynamics are systematically derived from first principles, are self-consistent \citep{strang_how_2019}, much more natural \citep{black_stochastic_2012}, and can fundamentally differ from the predictions made by simply adding noise terms to a deterministic model \citep{strang_how_2019}. Formulating the fundamental formal structures of evolutionary biology in terms of the mechanistic demographic processes of birth and death at the individual level is also greatly desirable for biological reasons \citep{metcalf_why_2007,geritz_mathematical_2012} because `all paths to fitness lead through demography' \citep{metcalf_all_2007}. In other words, since demographic processes such as birth and death rates explicitly account for the ecology of the system, they can more accurately reflect the complex interplay between ecological and evolutionary processes and provide a more fundamental mechanistic description of the relevant evolutionary forces and population dynamics \citep{doebeli_towards_2017}.

@@

In this thesis, I present a formulation of population dynamics constructed from mechanistic first principles grounded in individual-level birth and death. Part \ref{part_theory} presents the mathematical formalism in a detailed, self-contained, pedagogical manner. To facilitate readership by a broad audience, I only assume passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability. Familiarity with stochastic calculus is helpful for some sections but is not required. I present a brief introduction to the relevant mathematics in section \ref{sec_math_background} and present a toy example of tracking population size of a population of identical individuals in section \ref{sec_1D_processes}. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process, thus illustrating all the major tools required. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, use the system size expansion to arrive at a continuous description of change in trait frequencies as an SDE under mild assumptions on the functional forms of the birth and death rates. Unlike many classic stochastic formulations in evolutionary theory \citep{fisher_genetical_1930,wright_evolution_1931,moran_random_1958,crow_introduction_1970, lande_natural_1976}, I do not assume a fixed (effective) population size and instead allow the total population size to fluctuate over time. I show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, the Price equation), thus establishing the microscopic basis of well-known equations from stochastic first principles. I also illustrate some general predictions that can be made using the weak noise approximation for the sake of completeness. While the mathematics of section \ref{sec_nD_processes} is standard and well-understood, it has, to the best of my knowledge, not been used in this context in the generality we use here. Several specific models of specific systems do use these mathematical techniques, but these papers are often written assuming familiarity with notions in physics and/or mathematics and thus may not be very accessible to theoretical ecologists who do not have formal training in these subjects (but see \cite{czuppon_understanding_2021} for a recent pedagogical review on the general approach as applied to Wright-Fisher and Moran processes, where total population size is constant). As such, chapter \ref{sec_nD_processes} can be thought of as a tutorial, conceptual synthesis, and unification. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck in terms of functional derivatives. Under mild assumptions, I show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. I also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation. Chapter \ref{chap_infD_processes} generalizes the work of Tim Rogers and colleagues \citep{rogers_demographic_2012,rogers_spontaneous_2012,rogers_modes_2015}, and to the best of my knowledge, is entirely original both mathematically and biologically. Mathematically, the ideas present heuristic, accessible alternatives to the rigorous tools of martingale theory and measure-valued branching processes that are usually employed to describe the evolution of quantitative traits \citep{champagnat_unifying_2006,etheridge_mathematical_2011, week_white_2021} in terms of a `stochastic field theory'. This formulation in terms of stochastic field equations also opens up the model to analysis using powerful heuristic tools in physics such as the path integral \citep{doi_second_1976, peliti_path_1985, dodd_many-body_2009, chow_path_2015, weber_master_2017}, though I do not attempt to leverage these connections here, leaving it as an opportunity for future research. Examples of one-dimensional, multi-dimensional, and quantitative trait models that can be analyzed in this formalism are presented in Appendix \ref{App_examples}.\\
